### 2021-09-25 plos_latex_template.tex
update date
AST: addressing some of Liz's comments and tiny changes

Investing more in effort in the most excluded groups does not mean neglecting the others (see \textbf{Rule \ref{rule_inclusion}} for more on that). 
But it will prevent from falling into the mistake of treating diversity as a checklist.
It will also guide the vision of diversity for your conference--and your strategies to achieve it:
Will a more diverse conference would translate into an even gender distribution in your speakers? 
Would it be representation of all races--especially Black people--among the head organizers, speakers, and attendees? 
Would it be having LGBTQIA+ friendly-spaces or community participation from key geographic regions?
This exercise will also help coming up with indicators that would help the organizing team assess the fulfillment of the diversity objectives along the way.

@@

The following advice should apply to any kind of team, but are specially relevant when working with people from marginalized groups. First, rethink power relationships inside the group and share power. Be committed to examine power relations inside your organizing team and challenge it during all the process. % Rocío: Could we explain power relationship / share power ideas in more simple terms?
Don't expect self-nomination and voting to work as mechanisms to counteract systemic inequalities. Nominate directly and offer leading positions, and let people from privileged groups step down. 
Build an environment in which every person has their voice and enforce voices from systematically excluded groups.
% Liz: Use of "voice" in this way makes deaf people feel excluded. Maybe it could be something like "Build an environment in which every person has input..." Or anything as long as it doesn't suggest you have to have a voice to participate.
Offer mentorship and guidance if you encounter cases of impostor syndrome, doubts about the use of English in communication (See \textbf{Rule \ref{rule_language}}), or others. You can, for example, break the expectations about leadership as a lonely task and create smaller, co-led groups, where everyone finds their preferred tasks and gets to take leadership. Splitting the workload and responsibilities should not be done by putting care-taking labors--community building, meeting organization and note taking, conversations with potential partners--on the hands of women and other minoritized groups, while people from privileged groups take the lead in stereotypical highly-valued tasks (see \textbf{Rule \ref{rule_unbias}}). Make work visible, discuss collectively what should be done (and why), divide and account work. Being transparent and sharing the information is key to make work visible and collective. %sara: this relates to make labor visible from DF, maybe we can cite them here too. %Rocío: sure.
% Rocío: Idea of this paragraph: tips to make a diverse and inclusive team work.

% Liz: I have a little concern about mentoring WRT English, but I'm a white English speaker, so take it with a grain of salt. I think people who look like me need to do more accepting and less mentoring around English. If someone explicitly wants my help, I'm happy to give it because I love words and language, and I might learn something cool about another language. But I'm not sure we should be telling the English-priveleged readers that they are language mentors.

Most importantly, take care of your team. 
Having a diverse team and executing inclusive and accessible practices throughout the organizing period--it might be a year or two--may require a lot of effort. The effort is worth it because it strengthens the event and the community making it truly welcoming for everyone. 
However, having a strong community as the only reward may be enough for those who are in more privileged position. 
Some people and often the minoritized ones do not have institutional support to put time and effort into the organization tasks and do not have the luxury to commit to the organization for free. 
% Liz: Yes! This is especially true when *some* people are paid to work on the conference. I was just asked to do accessibility work that should have been done by the contracted, paid organizer. This is hugely problematic and will be true for lots of conferences outside the open software community.
In addition, tasks such as receiving and responding to code of conduct reports, can be emotionally intense work and should be additionally rewarded.
Prioritize your team's well-being. Check on them regularly and make sure everyone is comfortable. 
Be mindful of each particular context, be flexible with hours or commitments, and revise your budget (see \textbf{Rule \ref{rule_financial}}). 
% Rocío: Idea of this paragraph: take care of your team.
% Liz: also, include people from diverse backgrounds in all aspects of planning, not just the diversity-related ones, unless that is their only interest. In some organizations, I could be on the scientific committee and participate in many other ways, but I'm only called on for accessibility.

@@

Make sure that every selection committee--the committee looking for keynote speakers, the selection committee for abstracts, the prizes and award committees--are also diverse \cite{swartzScienceValueDiversity2019, wongBuildDiversityScience2020}, and ask them to be aware that everyone has implicit biases, to recognize them, and try to counteract them. 
An inclusive and diverse organizing team (\textbf{Rule \ref{rule_organizing_team}}) is already a great starting point to overcome this bias in other roles. The regional and local communities in your field are also good sources to tap into. 
For example, for useR!2021, groups like AfricaR (africa-r.org), R-Ladies (rladies.org), MiR (mircommunity.com), Forwards (forwards.github.io), and LatinR (latin-r.com) were fundamental to reach people for the organizing team, potential presenters, to co-lead social events, and attendees; some of them even held spaces to help the members of their groups to prepare abstracts for submission.
When organizing an international academic conference, you may reach local associations, student organizations, or other Early Career Researchers groups or communities of practice: ask for their requirements and barriers, and explicitly invite them to contribute to the conference. % Rocío: ask about their requirements and barriers sounds a bit confusing.
% Rocío: Not everyone knows what a community of practice is. Could we add a definition or use other words? Definition: "Communities of practice are groups of people who share a concern, a set of problems, or a passion about a topic, and who deepen their knowledge and expertise in this area by interacting on an ongoing basis". (Wenger, E., McDermott, R. A., & Snyder, W. (2002). Cultivating Communities of Practice: A Guide to Managing Knowledge. Harvard Business Press.)
Since many conferences are held with the primary goal of displaying cutting-edge work led by senior scientists and academics, early career researchers are often disadvantaged. Diversity should spread among all speaking roles, and the conference should also help raise the profiles of early career researchers, fostering collaboration and building their skills.
% Liz: maybe consider a broader group than "early career researchers." One problem is that marginalized people have many fewer opportunities, and as they age that attenuation of opportunities compounds. Someone can be in the margins even years after finishing grad school or being a postdoc (not to mention that people without degrees could have contributions.

@@

Networking and socializing in online conferences may be challenging, mostly because we are used to personal interactions at coffee or lunch breaks, or other in-person social settings \cite{salibaGettingGripsOnline2020}. 
However, social events in in-person conferences can be exclusive (e.g. galas or dinner nights at expensive venues), and they are a common place for code of conduct violations to occur. % Rocío: any reference to back the last statement? The CoC guide?
Organizers should invest time in creating opportunities to meet and bond virtually. 
An online event is a great opportunity to create spaces for people who do not enjoy in person networking. 
Some ideas in this line are: offering the option of written chat only, instead of voice or video conversations, opening events with team work like trivia, offer some events that can be passively enjoyed like movies, yoga sessions, or art displays, where attendants can choose to just sit and enjoy without talking, or have a chat channel to comment on their experiences during the session. Respect people's limits, preferences, and remember that `the usual' does not necessarily work with everyone in your community, and that one single networking activity will never serve the whole community. It is worth trying new and varied activities that might work with subgroups of attendees.
Moreover, virtual communication may make people from minoritized groups feel more included, thus participate more (e.g. \cite{trianaDoesOrderFacetoFace2012}).
% Rocío: Networking paragraph.

@@

% Rocío: Could start with an introduction (or put somewhere else) that online can easily make conferences more inclusive for many people, but not necessarily for people with disabilities, and that they can be easily left behind unless we actively work on their inclusion. 
Conferences are among the least accessible spaces that people with disabilities may encounter \cite{priceAccessImaginedConstruction2009}.
Ironically, accessibility practices are inclusive not only for people with disabilities but can be beneficial to a broad spectrum of people. For instance, having captions is helpful to deaf and hard-of-hearing people, non-native speakers, and everyone in general. 
% Liz: I don't think it's actually ironic. maybe surprising or counterintuitive?

If the conference is in person, the venue must be accessible for people with mobility limitations. 
$ Liz: let's stay away from deficit-model language like "limitations." Possible aternatives:
% people who use wheelchairs or other assistive mobility devices
% ... must be physically accessible for disabled people

@@

Presenters should always speak into a microphone to make it easier for the hard of hearing and for captioners or interpreters to listen to them. 
% Liz: avoid using disabilities as nouns like -- "the blind", "the deaf" etc. I woul duse "people who are deaf or hard-of-hearing" above. There's controversy about whether the word person
% should come first ("person who is blind") or the identity should come first ("blind person:) but the person should always be acknowledged (like "participants with disabilities", blind scientist", etc.. It's kind of objectifying without a person-related word.
Regardless of the conference format --online, in person, or hybrid --images used in communications about the conference should be accompanied by alternative text, while videos should have both captions and sound. 
% Liz: not just alt-text, the whole site should be screen-reader accessible so:
% communication about the conference should be fully screen-reader accessible including alt-text, and any videos should have both captions and sound.

@@

Most importantly, accessibility practices are not afterthoughts that can be dealt with at the last minute. 
They require time and early decision-making \cite{irishIncreasingParticipationUsing2020}. 
Conversely, inaccessible decisions are hard to course-correct, e.g., when finding out too late that a venue is inaccessible for wheelchairs in an in-person conference. 
% Liz: "for people who use whellchairs" rather than "for wheelchairs"

@@

% Rocío: Pasting Batool's text in her 10 simple rules because there are very valuable things we could use: "Rule: Meeting the need of linguistically diverse audience. Most international conferences are held in English. However, English is only spoken by 4.83 percent of the world population. If the panellist or speakers will not be presenting in their native language, it may helpful to rephrase audience questions in a less complex language. Consider letting speakers know in advance that they can ask for a question to be translated to them in their native language and that they can respond in their native language.    If possible, ensure that some sessions are delivered by languages other than English"

@@

\begin{itemize}
    \item Promotion: Actively reach out and promote the conference to people who have been systematically excluded/Outreach/promotion plan that reaches communities that are traditionally excluded in scientific meetings
% Liz: including explicit descriptions of accessibility and other inclusion practices. Help people answer the question "Can I attend this conference?" without the emotional labor of contacting the organizers.
    \item Express this welcoming spirit in your communication strategy: Be public about your commitment to equity and inclusion: display the CoC prominently in conference spaces, publish the diversity statement, fee waivers, abstract selection process
    \item Languages
    \item Words matter: Be careful how you talk about minoritized people, Assertively set tone of inclusion
    \item coherence between actions and words
    \item \textbf{networking} when you gather a diverse groups of attendants you also have to take care of the spaces where they will interact
\end{itemize}

If part of the community has been historically discriminated against, one should emphasize that they are particularly welcome in this event and that the organization will make it a safe and inclusive environment. Publish the diversity statement (e.g., \url{https://user2021.r-project.org/about/diversity_statement/}), the code of conduct and accessibility guidelines. Fee waivers, financial support, and ease of application/registration should also be advertised; people who fear getting rejected when applying for scholarships and waivers may find it relieving that the process will be supportive rather than discriminatory. Advertise the conference in multiple languages (rule \ref{rule_language}). Express this welcoming spirit in your communication strategy (social networks, website, brand, and visual identity) to let people know that they are seen, respected, and welcome; that this is their space and community too. 
% Yani: We need to work a little more here, perhaps with some examples.
% Rocío: could add some ideas from here 
% https://www.science.org.au/files/userfiles/support/emcr/documents/one-page-summary-emcr-improving-diversity-web.pdf
% Batool: I will work on this rule to include more details by 19th of Sept

@@

%sara: this fits better in the communication rule
Set the tone. %adding here from Liz's rule but maybe this could go in communication in general
In addition to excluding derogatory or discriminatory language, make the effort to teach yourself the vocabulary and the best ways to communicate to account for every culture and situation. Do not expect people to teach you --it's not their role-- and accept feedback without being defensive.
$ Liz: should be "without being offended"
\section{Allocate financial resources to support the presence of diverse people in every space of the conference
}
\label{rule_financial}

>>>>

Investing more in effort in the most excluded groups does not mean neglecting the others (see \textbf{Rule \ref{rule_inclusion}} for more on that). 
But it will prevent from falling into the mistake of treating diversity as a checklist.
It will also guide the vision of diversity for your conference--and your strategies to achieve it:
Will a more diverse conference translate into an even gender distribution in your speakers? 
Would it be the presence of racialized people --especially Black people-- among the head organizers, speakers, and attendees? 
Would it be having LGBTQIA+ friendly-spaces or community participation from key geographic regions?
This exercise will also help coming up with indicators that would help the organizing team assess the fulfillment of the diversity objectives along the way.

@@

The following advice should apply to any kind of team, but are specially relevant when working with people from marginalized groups. First, rethink power relationships inside the group and share power. Be committed to examine power relations inside your organizing team and challenge it during all the process. % Rocío: Could we explain power relationship / share power ideas in more simple terms?
Don't expect self-nomination and voting to work as mechanisms to counteract systemic inequalities. Nominate directly and offer leading positions, and let people from privileged groups step down. 
Build an environment in which every person can express their position and give prevalence to the positions from systematically excluded groups.
% Liz: Use of "voice" in this way makes deaf people feel excluded. Maybe it could be something like "Build an environment in which every person has input..." Or anything as long as it doesn't suggest you have to have a voice to participate.
%i rephrased slightly, avoiding "support" it's not a question of privileged people giving support but of positions being given importance
Offer support and guidance if you encounter cases of impostor syndrome, doubts about the use of English in communication (See \textbf{Rule \ref{rule_language}}), or others. You can, for example, break the expectations about leadership as a lonely task and create smaller, co-led groups, where everyone finds their preferred tasks and gets to take leadership. Splitting the workload and responsibilities should not be done by putting care-taking labors--community building, meeting organization and note taking, conversations with potential partners--on the hands of women and other minoritized groups, while people from privileged groups take the lead in stereotypical highly-valued tasks (see \textbf{Rule \ref{rule_unbias}}). Make work visible, discuss collectively what should be done (and why), divide and account work. Being transparent and sharing the information is key to make work visible and collective. %sara: this relates to make labor visible from DF, maybe we can cite them here too. %Rocío: sure.
% Rocío: Idea of this paragraph: tips to make a diverse and inclusive team work.

% Liz: I have a little concern about mentoring WRT English, but I'm a white English speaker, so take it with a grain of salt. I think people who look like me need to do more accepting and less mentoring around English. If someone explicitly wants my help, I'm happy to give it because I love words and language, and I might learn something cool about another language. But I'm not sure we should be telling the English-priveleged readers that they are language mentors.
%ast. i agree with liz and wrote this from the point of view of a non-native speaker- in terms of support rather than mentorship. replaced with support but feel free to check
Most importantly, take care of your team. 
Having a diverse team and executing inclusive and accessible practices throughout the organizing period--it might be a year or two--may require a lot of effort. The effort is worth it because it strengthens the event and the community making it truly welcoming for everyone. 
However, having a strong community as the only reward may be enough for those who are in more privileged position. 
Some people and often the minoritized ones do not have institutional support to put time and effort into the organization tasks and do not have the luxury to commit to the organization for free. 
% Liz: Yes! This is especially true when *some* people are paid to work on the conference. I was just asked to do accessibility work that should have been done by the contracted, paid organizer. This is hugely problematic and will be true for lots of conferences outside the open software community.
In addition, tasks such as receiving and responding to code of conduct reports, can be emotionally intense work and should be additionally rewarded.
Prioritize your team's well-being. Check on them regularly and make sure everyone is comfortable. 
Be mindful of each particular context, be flexible with hours or commitments, and revise your budget (see \textbf{Rule \ref{rule_financial}}). 
% Rocío: Idea of this paragraph: take care of your team.
% Liz: also, include people from diverse backgrounds in all aspects of planning, not just the diversity-related ones, unless that is their only interest. In some organizations, I could be on the scientific committee and participate in many other ways, but I'm only called on for accessibility.

@@

Make sure that every selection committee--the committee looking for keynote speakers, the selection committee for abstracts, the prizes and award committees--are also diverse \cite{swartzScienceValueDiversity2019, wongBuildDiversityScience2020}, and ask them to be aware that everyone has implicit biases, to recognize them, and try to counteract them. 
An inclusive and diverse organizing team (\textbf{Rule \ref{rule_organizing_team}}) is already a great starting point to overcome this bias in other roles. The regional and local communities in your field are also good sources to tap into. 
For example, for useR!2021, groups like AfricaR (africa-r.org), R-Ladies (rladies.org), MiR (mircommunity.com), Forwards (forwards.github.io), and LatinR (latin-r.com) were fundamental to reach people for the organizing team, potential presenters, to co-lead social events, and attendees; some of them even held spaces to help the members of their groups to prepare abstracts for submission.
When organizing an international academic conference, you may reach local associations, student organizations, or other Early Career Researchers groups or communities of practice: ask for their barriers for participation, and explicitly invite them to contribute to the conference. % Rocío: ask about their requirements and barriers sounds a bit confusing.
% Rocío: Not everyone knows what a community of practice is. Could we add a definition or use other words? Definition: "Communities of practice are groups of people who share a concern, a set of problems, or a passion about a topic, and who deepen their knowledge and expertise in this area by interacting on an ongoing basis". (Wenger, E., McDermott, R. A., & Snyder, W. (2002). Cultivating Communities of Practice: A Guide to Managing Knowledge. Harvard Business Press.)
Since many conferences are held with the primary goal of displaying cutting-edge work led by senior scientists and academics, early career researchers are often disadvantaged. Diversity should spread among all speaking roles, and the conference should also help raise the profiles of early career researchers and people from minoritized groups, fostering collaboration and building their skills.
% Liz: maybe consider a broader group than "early career researchers." One problem is that marginalized people have many fewer opportunities, and as they age that attenuation of opportunities compounds. Someone can be in the margins even years after finishing grad school or being a postdoc (not to mention that people without degrees could have contributions.

@@

Networking and socializing in online conferences may be challenging, mostly because we are used to personal interactions at coffee or lunch breaks, or other in-person social settings \cite{salibaGettingGripsOnline2020}. 
However, social events in in-person conferences can be exclusive (e.g. galas or dinner nights at expensive venues), and they are a common place for code of conduct violations to occur \cite{auroraHowRespondCode2019}. % Rocío: any reference to back the last statement? The CoC guide? ast: yes
Organizers should invest time in creating opportunities to meet and bond virtually. 
An online event is a great opportunity to create spaces for people who do not enjoy in person networking. 
Some ideas in this line are: offering the option of written chat only, instead of voice or video conversations, opening events with team work like trivia, offer some events that can be passively enjoyed like movies, yoga sessions, or art displays, where attendants can choose to just sit and enjoy without talking, or have a chat channel to comment on their experiences during the session. Respect people's limits, preferences, and remember that `the usual' does not necessarily work with everyone in your community, and that one single networking activity will never serve the whole community. It is worth trying new and varied activities that might work with subgroups of attendees.
Moreover, virtual communication may make people from minoritized groups feel more included, thus participate more (e.g. \cite{trianaDoesOrderFacetoFace2012}).
% Rocío: Networking paragraph.

@@

% Rocío: Could start with an introduction (or put somewhere else) that online can easily make conferences more inclusive for many people, but not necessarily for people with disabilities, and that they can be easily left behind unless we actively work on their inclusion. 
Conferences are among the least accessible spaces that people with disabilities may encounter \cite{priceAccessImaginedConstruction2009}.
Importantly, accessibility practices are inclusive not only for people with disabilities but can be beneficial to a broad spectrum of people. For instance, having captions is helpful to deaf and hard-of-hearing people, non-native speakers, and everyone in general. 
% Liz: I don't think it's actually ironic. maybe surprising or counterintuitive? ast: absolutely! i put importantly at the end to highlight that without opposition

If the conference is in person, the venue must be accessible for people with mobility limitations. 
% Liz: let's stay away from deficit-model language like "limitations." Possible aternatives:
% people who use wheelchairs or other assistive mobility devices
% ... must be physically accessible for disabled people

@@

Presenters should always speak into a microphone to make it easier for people who are hard of hearing and for captioners or interpreters to listen to them. 
% Liz: avoid using disabilities as nouns like -- "the blind", "the deaf" etc. I woul duse "people who are deaf or hard-of-hearing" above. There's controversy about whether the word person
% should come first ("person who is blind") or the identity should come first ("blind person:) but the person should always be acknowledged (like "participants with disabilities", blind scientist", etc.. It's kind of objectifying without a person-related word.
Regardless of the conference format --online, in person, or hybrid --images used in communications about the conference should be accompanied by alternative text, while videos should have both captions and sound. 
% Liz: not just alt-text, the whole site should be screen-reader accessible so:
% communication about the conference should be fully screen-reader accessible including alt-text, and any videos should have both captions and sound.

@@

Most importantly, accessibility practices are not afterthoughts that can be dealt with at the last minute. 
They require time and early decision-making \cite{irishIncreasingParticipationUsing2020}. 
Conversely, inaccessible decisions are hard to course-correct, e.g., when finding out too late that a venue is inaccessible for people who use wheelchairs in an in-person conference. 
% Liz: "for people who use whellchairs" rather than "for wheelchairs"

@@

% Rocío: Pasting Batool's text in her 10 simple rules because there are very valuable things we could use: "Rule: Meeting the need of linguistically diverse audience. Most international conferences are held in English. However, English is only spoken by 4.83 percent of the world population. If the panelist or speakers will not be presenting in their native language, it may helpful to rephrase audience questions in a less complex language. Consider letting speakers know in advance that they can ask for a question to be translated to them in their native language and that they can respond in their native language.    If possible, ensure that some sessions are delivered by languages other than English"

@@

\begin{itemize}
    \item Promotion: Actively reach out and promote the conference to people who have been systematically excluded/Outreach/promotion plan that reaches communities that are traditionally excluded in scientific meetings
    \item Include explicit descriptions of accessibility and other inclusion practices in your communication strategy. Help people answer the question "Can I attend this conference?" without the emotional labor of contacting the organizers.
    \item Express this welcoming spirit in your communication strategy: Be public about your commitment to equity and inclusion: display the CoC prominently in conference spaces, publish the diversity statement, fee waivers, abstract selection process
    \item Languages
    \item Words matter: Be careful how you talk about minoritized people, Assertively set tone of inclusion
    \item coherence between actions and words
    \item \textbf{networking} when you gather a diverse groups of attendants you also have to take care of the spaces where they will interact
\end{itemize}

If part of the community has been historically discriminated against, one should emphasize that they are particularly welcome in this event and that the organization will make it a safe and inclusive environment. Publish the diversity statement (e.g., \url{https://user2021.r-project.org/about/diversity_statement/}), the code of conduct and accessibility guidelines. Fee waivers, financial support, and ease of application/registration should also be advertised; people who fear getting rejected when applying for scholarships and waivers may find it relieving that the process will be supportive rather than discriminatory. Advertise the conference in multiple languages (rule \ref{rule_language}). Express this welcoming spirit in your communication strategy (social networks, website, brand, and visual identity) to let people know that they are seen, respected, and welcome; that this is their space and community too. 
% Yani: We need to work a little more here, perhaps with some examples.
% Rocío: could add some ideas from here 
% https://www.science.org.au/files/userfiles/support/emcr/documents/one-page-summary-emcr-improving-diversity-web.pdf
% Batool: I will work on this rule to include more details by 25th of Sept

@@

%sara: this fits better in the communication rule
Set the tone. %adding here from Liz's rule but maybe this could go in communication in general
In addition to excluding derogatory or discriminatory language, make the effort to teach yourself the vocabulary and the best ways to communicate to account for every culture and situation. Do not expect people to teach you --it's not their role-- and accept feedback without being offended.

\section{Allocate financial resources to support the presence of diverse people in every space of the conference
}
\label{rule_financial}
